---
title: "R Notebook"
author: thread-0
date: 8/20/20
output: html_notebook
---

The [R plugin](https://www.jetbrains.com/help/pycharm/r-plugin-support.html) for IntelliJ-based IDEs provides
handy capabilities to work with the [R Markdown](https://www.jetbrains.com/help/pycharm/r-markdown.html) files.
To [add](https://www.jetbrains.com/help/pycharm/r-markdown.html#add-code-chunk) a new R chunk,
position the caret at any line or the code chunk, then click "+".

The code chunk appears:
```{r}
library(sparklyr)
library(dplyr)
library(microbenchmark)
```

```{r}
sc <- spark_connect(master = "spark://master:7077")
# sc <- spark_connect(master = "local")
###, config = list("sparklyr.shell.driver-memory" = "6g"))
data <- data.frame(y = runif(10^3, 0, 1))

## Benchmark Data copy_to()
microbenchmark::microbenchmark(
  setup = library(arrow),
  arrow_on = {
    sparklyr_df <<- copy_to(sc, data, overwrite = T)
    count(sparklyr_df) %>% collect()
  },
  arrow_off = {
    if ("arrow" %in% .packages()) detach("package:arrow")
    sparklyr_df <<- copy_to(sc, data, overwrite = T)
    count(sparklyr_df) %>% collect()
  },
  times = 10
) %T>% print() %>% ggplot2::autoplot()

spark_disconnect(sc)
```

```{r}
# sc <- spark_connect(master = "local")
sc <- spark_connect(master = "spark://master:7077")

# Benchmark collect()
microbenchmark::microbenchmark(
  setup = library(arrow),
  arrow_on = {
    collect(sparklyr_df)
  },
  arrow_off = {
    if ("arrow" %in% .packages()) detach("package:arrow")
    collect(sparklyr_df)
  },
  times = 10
) %T>% print() %>% ggplot2::autoplot()
spark_disconnect(sc)
```

```{r}
# sc <- spark_connect(master = "local")
sc <- spark_connect(master = "spark://master:7077")

## Benchmark data transformation
microbenchmark::microbenchmark(
  setup = library(arrow),
  arrow_on = {
    sample_n(sparklyr_df, 10^2) %>% spark_apply(~ .x / 2) %>% count()
  },
  arrow_off = {
    if ("arrow" %in% .packages()) detach("package:arrow")
    sample_n(sparklyr_df, 10^2) %>% spark_apply(~ .x / 2) %>% count()
  },
  times = 10
) %T>% print() %>% ggplot2::autoplot()
spark_disconnect(sc)
```
Type any R code in the chunk, for example:
```{r}
mycars <- within(mtcars, { cyl <- ordered(cyl) })
mycars
```

Now, click the **Run** button on the chunk toolbar to [execute](https://www.jetbrains.com/help/pycharm/r-markdown.html#run-r-code) the chunk code. The result should be placed under the chunk.
Click the **Knit and Open Document** to built and preview an output.