---
title: "NISS Overview"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```
Day 1 8:30–10:00
```
# R Fundamentals (N1)

## Data Science Workflows

## RStudio

## Data Structures

## Functions

## Functional Programming

```
10:30–12:00
```
# Data Sources (N2)

## Plain Text Files

## JSON

## Spreadsheets

## Databases (PostgreSQL)

## Web Scraping

```
1:00–2:30
```
# Data Transformation (N3)

## Data Cleaning (tidyr)

## Data Manipulation (dplyr)

```
3:00–4:30 
```
# HDFS/ Hadoop (N4)

## HDFS (rhdfs)

```{r}
options(warn=-1)
library(rhdfs)
hdfs.init()
```

## Data warehousing with Hive

## MapReduce (rmr2)

```{r}
groups = rbinom(32, n = 200, prob = 0.4)
head(groups)
tapply(groups, groups, length)
```

```{r}
library(rmr2)
groups = to.dfs(groups)
hdfs.ls(path="/user/rstudio")
out.data <- from.dfs(
  mapreduce(
    input = groups, 
    map = function(., v) keyval(v, 1), 
    reduce = function(k, vv)  keyval(k, length(vv))))
out.data
plot(out.data$key, out.data$val/200, type = "h", lwd=3, xlab = "y", ylab = "p(y)")
```

```{r}
options(warn=0)
```

## R Scripts

```
5:00–6:00 
```
# Introduction to Spark (N5)

## Spark Basics

## R Interfaces to Spark

```
Day 2 8:30–10:00
```
# Spark DataFrames (N6)

## RStudio's dplyr and sparklyr interfaces

## Spark DaraFrame manipulation with dplyr

## Spark’s SparkR interface

## Spark DataFrame creation and manipulation using SparkR

## Running SQL queries from SparkR

```
10:30–12:00 
```
# Supervised Learning (N7)

## Overview of models

## Regression: linear least squares, Lasso, ridge regression, decision trees, random forests, gradient-boosted trees

## Classification: logistic regression, decision trees, random forests, gradient-boosted trees, naive Bayes

## sparklyr and SparkR interfaces to these models.

```
1:00–2:30 
```
# Unsupervised Learning (N8)

## Overview of methods

## Clustering: k-means

## Dimension reduction: PCA

## sparklyr and SparkR interfaces to these methods.

```
3:00–4:30
```
# H2O and Other Extensions (N9)

##Overview of models

## Gradient Boosting Machine

## Deep Learning

```{r}
library(h2o)
h2o.init(nthreads=-1, max_mem_size="2G")
h2o.removeAll()

# args(h2o.deeplearning)
# help(h2o.deeplearning)
example(h2o.deeplearning)
# demo(h2o.deeplearning)
h2o.shutdown(prompt=FALSE)
```

## Grid Search

## rsparkling interfaces to these models

```
5:00–6:00 
```
## In-depth Examples (N10)

## Wrap-up